---
title: "Differential expression analysis of individual Drosophila melanogaster (w1118) male and female flies fed with different concentrations of dimethyl sulfoxide (DMSO) over different time periods."
author: "Analysis done by Yuejun Wang, TriLab Biounformatic Group"
output:
  html_document:
  pdf_document: default
  df_print: paged
gemoetry: margin=1in
fontsize: 8pt
fontfamily: mathpazo
---

\center

# *Summary:*
To gain an understanding of the toxic effect of a commonly used solvent, flies were exposed to 0, 0.5, 2, 3.5, or 7% v/v DMSO for 2, 4, 8, 24 or 48 h. We performed compound exposure of 800 individual flies in 4 Whole Animal Feeding Flats (WAFFL), a novel 96 well system to house, feed, and harvest individual flies. This expression profiling was part of a set of the experiments performed to evaluate the suitability of the WAFFL for high throughput small compound screening in D. melanogaster. Treated flies and controls were used for poly A+ stranded mRNA library preparation and we performed high throughput RNA sequencing to determine the transcriptional changes due to DMSO treatment.

```{r setup, include=FALSE, message=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE, tidy = TRUE)
```

# Load libraries
```{r message=FALSE}

options(java.parameters = "-Xmx8000m")

library(tidyverse)
library(DESeq2)
library(cowplot)
library(ggpubr)
library(RColorBrewer)
library(pheatmap)
library(ggsci)
library(AnnotationDbi)
library(scales)
library(org.Dm.eg.db)
library(xlsx)
library(r2excel)
library(ggplot2)
library(viridis)
library(RSkittleBrewer)
library(genefilter)
library(corrplot)
library(reshape2)
library(ComplexHeatmap)
library(circlize)
library(UpSetR)
trop = RSkittleBrewer('tropical')
```

# Load aux functions
```{r}
source(file = "../01_aux_rnaseq_functions.R")
```

# Load data
```{r}
metadata <- as.data.frame(read_tsv(file = "../SampleInfo.txt", col_names = TRUE, comment = "#"))
rownames(metadata) <- metadata$SampleName


read_counts <- as.data.frame(read_tsv(file = "/gpfs/gsfs12/users/wangy80/TK74/results/05counts/read_counts", col_names = TRUE, comment = "#"))
rownames(read_counts) <- read_counts$Geneid # adding gene ids as row names


# Rename sample names
read_counts <- read_counts[,7:length(read_counts)]

colnames(read_counts) <- gsub(".sorted.dedup.bam","",
                              gsub("results/04dedup/","",
                                   colnames(read_counts)))


# Sort tables so metadata and read counts match order
read_counts <- read_counts[,match(metadata$SampleName, colnames(read_counts))]
all(rownames(metadata) == colnames(read_counts))


# Round read counts to the closest interger
read_counts <- round(read_counts, digits = 0)

# include total read counts in metadata
metadata$read_counts <- colSums(read_counts)

# Rename column names in read_counts based on metadata
colnames(read_counts) <- rownames(metadata)

write.table(x = metadata, file = "metadata.txt", sep = "\t") 
```

# DE analysis with DESeq2

```{r}
dir.create(path = "./Plots", showWarnings = FALSE)

# Converting Time, DMSO and Replicate number to factors
metadata$Time <- factor(metadata$Time,levels=c("2hr","4hr","8hr","24hr","48hr"))
metadata$DMSO <- as.factor(metadata$DMSO)
metadata$Replicate <- as.factor(metadata$Replicate)
metadata$Sex <- as.factor(metadata$Sex)

# Adding read_depth in design to control for read_depth
dds <- DESeqDataSetFromMatrix(countData = read_counts, 
                              colData = metadata,  
                              design = ~ Sex + Time + DMSO)

```


# Separate M and F

```{r}
# female
dds.f <- dds[, dds$Sex == "F"]

dds.f <- DESeqDataSetFromMatrix(countData = counts(dds.f), 
                              colData = colData(dds.f),  
                              design = ~ Time + DMSO)

# Normalize counts
vsd.f <- vst(dds.f, blind=FALSE)

# Keep genes with 10 reads in at least 3 samples
keep1 <- rowSums(counts(dds.f) >= 10) >= 3
vsd.f <- vsd.f[keep1,]
df.f <- as.data.frame(assay(vsd.f))

df.f$gene <- rownames(df.f)
df.f.m <- melt(df.f)

df.f.m$variable <- str_replace_all(df.f.m$variable, "M_0_8hr_rep2", "F_0_8hr_rep2")

df.f.m$sample <- gsub("_rep[1,2,3]","",df.f.m$variable)
df.f.m <- df.f.m[which(df.f.m$variable!="M_0_24hr_rep1"),]

write.table(df.f,"female_DESeq2_normalized.txt",sep="\t",quote=F)


# male
dds.m <- dds[, dds$Sex == "M"]

dds.m <- DESeqDataSetFromMatrix(countData = counts(dds.m), 
                              colData = colData(dds.m),  
                              design = ~ Time + DMSO)

# Normalize counts
vsd.m <- vst(dds.m, blind=FALSE)

# Keep genes with 10 reads in at least 3 samples
keep2 <- rowSums(counts(dds.m) >= 10) >= 3
vsd.m <- vsd.m[keep2,]
df.m <- as.data.frame(assay(vsd.m))

df.m$gene <- rownames(df.m)
df.m.m <- melt(df.m)
df.m.m$variable <- str_replace_all(df.m.m$variable, "F_0_8hr_rep2", "M_0_8hr_rep2")

df.m.m$sample <- gsub("_rep[1,2,3]","",df.m.m$variable)
df.m.m <- df.m.m[which(df.m.m$variable!="M_0_24hr_rep1"),]

write.table(df.m,"male_DESeq2_normalized.txt",sep="\t",quote=F)

```

# wilcox.test for specific group to all the other samples

```{r}
rank_test <- function(data){
  groups <- unique(data$sample)

  ## remove samples with only one replicate
  groups <- groups[-which(groups %in% c('F_3.5_24hr','M_3.5_48hr'))]
  df_list <- list()
  for(i in 1:length(groups)){
    res <- data %>%
      group_by(gene) %>%
      summarise(across(value, ~ wilcox.test(.[sample == groups[i]], .[sample != groups[i]],exact = FALSE)$p.value, .names="pvalue"),
              across(value, ~ mean(.[sample == groups[i]]),.names=groups[i]),
              across(value, ~ mean(.[sample != groups[i]]),.names=paste0("not_",groups[i]))
              ) %>%
      as.data.frame
  res$diff <- res[,3]-res[,4]
  res$p.adjust <- p.adjust(res$pvalue,method="fdr")
  
  colnames(res)[5] <- paste0(groups[i],".diff")
  colnames(res)[6] <- paste0(groups[i],".fdr")
  
  assign(groups[i],res)
  id <- groups[i]
  df_list[[id]] <- res[,c(1,5,6)]
  }
  
  return(df_list)
}

```


```{r}

df_list_male <- rank_test(data=df.m.m)
df_list_female <- rank_test(data=df.f.m)

```


```{r}
# Join multiple data.frames
# merge all data frames in list
out_male <- df_list_male %>% purrr::reduce(full_join, by='gene')
write.table(out_male,"male_comparison_wilcoxtest.txt",quote=F,sep="\t",row.names=F)

out_female <- df_list_female %>% purrr::reduce(full_join, by='gene')
write.table(out_female,"female_comparison_wilcoxtest.txt",quote=F,sep="\t",row.names=F)


```


# heatmap

```{r message=FALSE}
# rename column names
colnames <- c("gene","diff","fdr") 

# male
df_list_male2 <- lapply(df_list_male, setNames, colnames)

merged_df_male <- bind_rows(df_list_male2, .id = "comparisons")

df_wide_male <- pivot_wider(merged_df_male[,c(1,2,3)],  names_from = "comparisons", values_from = "diff")
mat_male <- as.data.frame(df_wide_male[,2:ncol(df_wide_male)])
rownames(mat_male) <- df_wide_male$gene

mat_male[is.na(mat_male)] <- 0

p1 <- Heatmap(mat_male, 
      name="diff",
      show_row_names = FALSE,
      col = colorRamp2(c(-1, 0, 1), c("blue", "white", "red"))
      #column_names_gp = gpar(fontsize =8),
      #heatmap_height = unit(12, "cm"), 
      #heatmap_width = unit(12, "cm")
      )

pdf(paste0("./Plots/male_pheatmap_","diff",".pdf"),p1,height=8,width=8)
plot(p1)
dev.off()

# female
df_list_female2 <- lapply(df_list_female, setNames, colnames)

merged_df_female <- bind_rows(df_list_female2, .id = "comparisons")

df_wide_female <- pivot_wider(merged_df_female[,c(1,2,3)],  names_from = "comparisons", values_from = "diff")
mat_female <- as.data.frame(df_wide_female[,2:ncol(df_wide_female)])
rownames(mat_female) <- df_wide_female$gene

mat_female[is.na(mat_female)] <- 0

p2 <- Heatmap(mat_female, 
      name="diff",
      show_row_names = FALSE,
      col = colorRamp2(c(-1, 0, 1), c("blue", "white", "red"))
      #column_names_gp = gpar(fontsize =8),
      #heatmap_height = unit(12, "cm"), 
      #heatmap_width = unit(12, "cm")
      )

pdf(paste0("./Plots/female_pheatmap_","diff",".pdf"),p2,height=8,width=8)
plot(p2)
dev.off()

```

```{r}
sessionInfo()
```
